Parameter Recovery Validation

### 2025-05-29 nbs/concepts/parameter-recovery-validation/index.qmd
docs: remove unused subsections
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

### Key Advantages of Gamma/Normal Hierarchical Time Structure

@@

#### **1. Activation Patterns**
**Characteristics**: Genes that start with low expression and become strongly activated during the process.

**Parameter specifications**:
- $\alpha^*_{\text{off}i} < 0.2$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.5$ (strong activation)
- $t^*_{0,\text{on}i} < 0.4$ (early activation onset)
- $\delta^*_i > 0.3$ (sustained activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 7.5$ (strong upregulation)

@@

#### **2. Decay-Only Patterns**
**Characteristics**: Genes that start with high expression and gradually decrease without reactivation.

**Parameter specifications**:
- $\alpha^*_{\text{off}i} > 0.5$ (high basal transcription)
- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window)
- $\delta^*_i$ (irrelevant, no activation observed)
- **Effective dynamics**: Monotonic decay from high initial levels

@@

**Parameter specifications**:
- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (moderate to strong activation)
- $t^*_{0,\text{on}i} < 0.5$ (early to mid-process activation)
- $\delta^*_i < 0.3$ (brief activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (clear but not extreme upregulation)

@@

**Parameter specifications**:
- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (strong activation)
- $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
- $\delta^*_i > 0.6$ (long activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (sustained upregulation)

@@

#### **Pattern Classification Criteria**

For parameter recovery validation, genes can be automatically classified using these decision rules:

```python
def classify_gene_pattern(alpha_off, alpha_on, t_on, delta, T_max):
    """Classify gene expression pattern based on inferred parameters."""
    fold_change = alpha_on / alpha_off

    if t_on > T_max:
        return "decay_only"
    elif alpha_off < 0.2 and alpha_on > 1.5 and t_on < 0.4 and delta > 0.3:
        return "activation"
    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.5 and delta < 0.3:
        return "transient"
    elif alpha_off < 0.3 and fold_change > 3.3 and t_on < 0.3 and delta > 0.6:
        return "sustained"
    else:
        return "intermediate"  # Mixed or boundary cases
```

#### **Validation Strategy for Pattern Recovery**

Parameter recovery validation should test the model's ability to:

1. **Correctly classify known patterns**: Generate synthetic data with known pattern types and verify classification accuracy
2. **Recover pattern-specific parameters**: Ensure fold-changes, timing, and duration parameters are accurately inferred
3. **Distinguish similar patterns**: Separate transient vs. sustained activation, activation vs. decay-only
4. **Handle boundary cases**: Genes with intermediate parameter values that don't clearly fit one pattern

@@

### Key Advantages of the Piecewise Framework

@@

### Implementation Considerations

#### **Phase Detection Logic**

For implementation, each cell-gene pair $(j,i)$ must be assigned to the correct phase based on the cell's latent time $t^*_j$ relative to gene $i$'s activation parameters:

```python
def determine_phase(t_star_j, t_on_i, delta_i):
    """Determine which phase cell j is in for gene i."""
    if t_star_j < t_on_i:
        return "off_phase_1"  # Phase 1: before activation
    elif t_star_j < t_on_i + delta_i:
        return "on_phase"     # Phase 2: during activation
    else:
        return "off_phase_3"  # Phase 3: after activation
```

#### **Numerical Stability Considerations**

**Special case handling for $\gamma^*_i \approx 1$:**

When $\gamma^*_i$ is very close to 1 (e.g., $|\gamma^*_i - 1| < 10^{-6}$), the helper variable $\xi_i = \frac{\alpha^*_{\text{off}i} - \alpha^*_{\text{on}i}}{\gamma^*_i - 1}$ becomes numerically unstable. Implementation should:

1. **Use threshold switching**: If $|\gamma^*_i - 1| < \epsilon$ (e.g., $\epsilon = 10^{-6}$), use the $\gamma^*_i = 1$ analytical solutions
2. **L'Hôpital's rule**: For intermediate values, the limit as $\gamma^*_i \to 1$ gives the correct $t^* e^{-t^*}$ terms
3. **Robust computation**: Avoid direct division by $(\gamma^*_i - 1)$ when close to zero

**Parameter validation ranges:**

To ensure biological realism and numerical stability:

```python
def validate_parameters(alpha_off, alpha_on, t_on, delta, gamma_star):
    """Validate parameter ranges for biological realism."""
    assert 0.01 <= alpha_off <= 10.0    # Reasonable basal transcription
    assert 0.1 <= alpha_on <= 20.0      # Reasonable active transcription
    assert alpha_on > alpha_off          # Activation increases transcription
    assert 0.01 <= t_on <= 0.9          # Onset within process timeline
    assert 0.05 <= delta <= 2.0         # Reasonable activation duration
    assert 0.1 <= gamma_star <= 10.0    # Reasonable relative degradation
```

@@

>>>>

### Properties of Gamma/Normal Hierarchical Time Structure

@@

#### **1. Activation Patterns**

**Characteristics**: Genes that start with low expression and become strongly activated during the process.

**Parameter specifications**:

- $\alpha^*_{\text{off}i} < 0.2$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.5$ (strong activation)
- $t^*_{0,\text{on}i} < 0.4$ (early activation onset)
- $\delta^*_i > 0.3$ (sustained activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 7.5$ (strong upregulation)

@@

#### **2. Decay-Only Patterns**

**Characteristics**: Genes that start with high expression and gradually decrease without reactivation.

**Parameter specifications**:

- $\alpha^*_{\text{off}i} > 0.5$ (high basal transcription)
- $t^*_{0,\text{on}i} > T^*_M$ (activation onset beyond observation window)
- $\delta^*_i$ (irrelevant, no activation observed)
- **Effective dynamics**: Monotonic decay from high initial levels

@@

**Parameter specifications**:

- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (moderate to strong activation)
- $t^*_{0,\text{on}i} < 0.5$ (early to mid-process activation)
- $\delta^*_i < 0.3$ (brief activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (clear but not extreme upregulation)

@@

**Parameter specifications**:

- $\alpha^*_{\text{off}i} < 0.3$ (low basal transcription)
- $\alpha^*_{\text{on}i} > 1.0$ (strong activation)
- $t^*_{0,\text{on}i} < 0.3$ (early activation onset)
- $\delta^*_i > 0.6$ (long activation duration)
- **Fold-change**: $\alpha^*_{\text{on}i}/\alpha^*_{\text{off}i} > 3.3$ (sustained upregulation)

@@

@@

### Properties of the Piecewise Framework

@@

@@

### Implementation and Validation Results

The piecewise activation model has been successfully implemented within the PyroVelocity modular framework. This section demonstrates the complete parameter recovery validation workflow, from model creation through comprehensive validation across all four gene expression patterns.

#### Model Creation and Setup

The piecewise activation model is created using the PyroVelocity factory system with specialized components for piecewise dynamics and hierarchical priors:

```python
import torch
import pyro
import numpy as np
import matplotlib.pyplot as plt
import seaborn as sns
from pathlib import Path

# PyroVelocity imports
from pyrovelocity.models.modular.factory import create_piecewise_activation_model
from pyrovelocity.plots.piecewise import (
    plot_piecewise_phase_portrait,
    plot_piecewise_trajectories,
    plot_activation_timing,
)
from pyrovelocity.plots.predictive_checks import (
    plot_prior_predictive_checks,
    plot_posterior_predictive_checks,
    plot_parameter_recovery_diagnostics,
)

# Create the piecewise activation model
model = create_piecewise_activation_model(
    num_genes=10,
    num_cells=200,
    name="piecewise_activation_study3"
)

print(f"Created model: {model}")
print(f"Model components:")
print(f"  Dynamics: {model.dynamics_model}")
print(f"  Prior: {model.prior_model}")
print(f"  Likelihood: {model.likelihood_model}")
print(f"  Observation: {model.observation_model}")
print(f"  Guide: {model.guide_model}")
```

#### Synthetic Data Generation for All Expression Patterns

We generate synthetic datasets representing each of the four gene expression patterns using the generic predictive sampling interface:

```python
# Generate validation datasets for all patterns using generic API
validation_datasets = {}
validation_parameters = {}

# Define pattern configurations (2 parameter sets per pattern = 8 total)
pattern_configs = [
    # Activation patterns
    {"pattern": "activation", "set_id": 1, "seed": 42},
    {"pattern": "activation", "set_id": 2, "seed": 43},
    # Decay patterns
    {"pattern": "decay", "set_id": 1, "seed": 44},
    {"pattern": "decay", "set_id": 2, "seed": 45},
    # Transient patterns
    {"pattern": "transient", "set_id": 1, "seed": 46},
    {"pattern": "transient", "set_id": 2, "seed": 47},
    # Sustained patterns
    {"pattern": "sustained", "set_id": 1, "seed": 48},
    {"pattern": "sustained", "set_id": 2, "seed": 49},
]

for config in pattern_configs:
    pattern = config["pattern"]
    set_id = config["set_id"]
    seed = config["seed"]
    dataset_key = f"{pattern}_{set_id}"

    print(f"Generating synthetic data for {dataset_key}...")

    # Set PyTorch random seed for reproducibility
    torch.manual_seed(seed)
    pyro.set_rng_seed(seed)

    # Sample system parameters constrained to this pattern
    true_system_params = model.sample_system_parameters(
        pattern=pattern,
        set_id=set_id,
        num_samples=1,
        constrain_to_pattern=True
    )

    # Generate synthetic data using the generic predictive sampling interface
    torch.manual_seed(seed + 100)
    pyro.set_rng_seed(seed + 100)
    adata = model.generate_predictive_samples(
        num_cells=200,
        num_genes=5,
        samples=true_system_params,  # Use constrained parameters as "posterior"
        return_format="anndata"
    )

    # Store dataset and true parameters
    validation_datasets[dataset_key] = adata
    validation_parameters[dataset_key] = true_system_params

    print(f"  Generated {adata.n_obs} cells × {adata.n_vars} genes")
    print(f"  Pattern: {adata.uns['pattern_type']}")
    print(f"  True parameters shape: {[f'{k}: {v.shape}' for k, v in true_system_params.items()]}")

print(f"\nGenerated {len(validation_datasets)} synthetic datasets")
```

#### Prior Predictive Checks

Before training, we validate that our priors generate biologically reasonable parameter ranges and expression patterns:

```python
# Generate prior predictive samples using the generic interface
print("Performing prior predictive checks...")

# Set seeds for reproducible prior sampling
torch.manual_seed(0)
pyro.set_rng_seed(0)

# Generate prior predictive samples (no samples argument = sample from prior)
prior_predictive_adata = model.generate_predictive_samples(
    num_cells=200,
    num_genes=10,
    num_samples=100,  # Generate 100 prior samples
    return_format="anndata"
)

# Extract parameter samples for validation
torch.manual_seed(1)
pyro.set_rng_seed(1)
prior_parameter_samples = model.sample_system_parameters(
    num_samples=1000,
    constrain_to_pattern=False  # Use full prior ranges
)

# Plot prior predictive checks
fig_prior = plot_prior_predictive_checks(
    model=model,
    prior_adata=prior_predictive_adata,
    prior_parameters=prior_parameter_samples,
    figsize=(15, 10)
)

plt.suptitle("Prior Predictive Checks: Piecewise Activation Model", fontsize=16)
plt.tight_layout()
plt.show()

# Validate prior parameter ranges
print("\nPrior parameter range validation:")
for param_name, samples in prior_parameter_samples.items():
    if param_name.startswith(('alpha_off', 'alpha_on', 't_on', 'delta', 'gamma_star')):
        print(f"{param_name}:")
        print(f"  Range: [{samples.min():.3f}, {samples.max():.3f}]")
        print(f"  Mean ± Std: {samples.mean():.3f} ± {samples.std():.3f}")

# Note: Pattern classification and other specialized validation methods
# will be implemented later as separate functionality
```

#### Parameter Recovery Validation Across All Patterns

We now perform comprehensive parameter recovery validation across all eight parameter sets using the built-in model validation method:

```python
# Run parameter recovery validation for all datasets using built-in model method
validation_results = {}
posterior_samples_all = {}

for dataset_key, adata in validation_datasets.items():
    print(f"\n{'='*60}")
    print(f"Validating parameter recovery for {dataset_key}")
    print(f"{'='*60}")

    # Create fresh model instance for this validation
    model_instance = create_piecewise_activation_model(
        num_genes=adata.n_vars,
        num_cells=adata.n_obs,
        name=f"piecewise_{dataset_key}"
    )

    # Run parameter recovery validation using built-in model method
    result = model_instance.validate_parameter_recovery(
        adata=adata,
        true_parameters=validation_parameters[dataset_key],
        max_epochs=2000,
        num_posterior_samples=500,
        convergence_threshold=0.95
    )

    validation_results[dataset_key] = result
    posterior_samples_all[dataset_key] = result["posterior_samples"]

    # Print summary results
    print(f"Parameter recovery summary for {dataset_key}:")
    print(f"  Overall success rate: {result['overall_success_rate']:.1%}")
    print(f"  Training converged: {result['training_result']['converged']}")
    print(f"  Final ELBO: {result['training_result']['final_elbo']:.2f}")

    # Print parameter-specific recovery rates
    for param_name, recovery_rate in result['recovery_metrics']['parameter_recovery_rates'].items():
        print(f"  {param_name}: {recovery_rate:.1%}")

# Compute overall validation statistics
overall_success_rates = [r['overall_success_rate'] for r in validation_results.values()]
mean_success_rate = np.mean(overall_success_rates)
std_success_rate = np.std(overall_success_rates)

print(f"\n{'='*60}")
print(f"OVERALL VALIDATION SUMMARY")
print(f"{'='*60}")
print(f"Mean parameter recovery rate: {mean_success_rate:.1%} ± {std_success_rate:.1%}")
print(f"Datasets with >90% recovery: {sum(r >= 0.9 for r in overall_success_rates)}/{len(overall_success_rates)}")
print(f"Datasets with >95% recovery: {sum(r >= 0.95 for r in overall_success_rates)}/{len(overall_success_rates)}")
```

#### Comprehensive Visualization Suite

We generate a comprehensive set of validation plots to visualize model performance:

```python
# Create visualization directory
viz_dir = Path("validation_plots")
viz_dir.mkdir(exist_ok=True)

print("Generating comprehensive validation visualizations...")

# 1. Parameter recovery diagnostic plots
fig_recovery = plot_parameter_recovery_diagnostics(
    true_parameters=validation_parameters,
    inferred_parameters={k: v["posterior_samples"] for k, v in validation_results.items()},
    figsize=(20, 15)
)
plt.suptitle("Parameter Recovery Diagnostics: All Patterns", fontsize=16)
plt.savefig(viz_dir / "parameter_recovery_diagnostics.png", dpi=300, bbox_inches='tight')
plt.show()

# 2. Phase portraits for each pattern
for pattern in ["activation", "decay", "transient", "sustained"]:
    # Find datasets for this pattern
    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}

    for dataset_key, adata in pattern_datasets.items():
        fig_phase = plot_piecewise_phase_portrait(
            adata=adata,
            gene_names=adata.var_names[:3].tolist(),  # Plot first 3 genes
            posterior_samples=posterior_samples_all[dataset_key],
            true_parameters=validation_parameters[dataset_key],
            figsize=(15, 10)
        )
        plt.suptitle(f"Phase Portrait: {dataset_key.replace('_', ' ').title()}", fontsize=16)
        plt.savefig(viz_dir / f"phase_portrait_{dataset_key}.png", dpi=300, bbox_inches='tight')
        plt.show()

# 3. Temporal trajectories showing piecewise dynamics
for pattern in ["activation", "transient", "sustained"]:  # Skip decay (no activation)
    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}

    for dataset_key, adata in pattern_datasets.items():
        fig_traj = plot_piecewise_trajectories(
            adata=adata,
            gene_names=adata.var_names[:2].tolist(),  # Plot first 2 genes
            posterior_samples=posterior_samples_all[dataset_key],
            time_points=np.linspace(0, 1, 100),
            figsize=(12, 8)
        )
        plt.suptitle(f"Temporal Trajectories: {dataset_key.replace('_', ' ').title()}", fontsize=16)
        plt.savefig(viz_dir / f"trajectories_{dataset_key}.png", dpi=300, bbox_inches='tight')
        plt.show()

# 4. Activation timing analysis
fig_timing = plot_activation_timing(
    posterior_samples=posterior_samples_all,
    true_parameters=validation_parameters,
    figsize=(15, 10)
)
plt.suptitle("Activation Timing Analysis: All Patterns", fontsize=16)
plt.savefig(viz_dir / "activation_timing_analysis.png", dpi=300, bbox_inches='tight')
plt.show()

# 5. Posterior predictive checks for each pattern
for pattern in ["activation", "decay", "transient", "sustained"]:
    pattern_datasets = {k: v for k, v in validation_datasets.items() if k.startswith(pattern)}

    for dataset_key, adata in pattern_datasets.items():
        fig_post = plot_posterior_predictive_checks(
            model=model,
            adata=adata,
            posterior_samples=posterior_samples_all[dataset_key],
            figsize=(15, 10)
        )
        plt.suptitle(f"Posterior Predictive Checks: {dataset_key.replace('_', ' ').title()}", fontsize=16)
        plt.savefig(viz_dir / f"posterior_predictive_{dataset_key}.png", dpi=300, bbox_inches='tight')
        plt.show()

print(f"All validation plots saved to {viz_dir}/")
```
